---
title: "How to get away with selection. Part I: Introduction"
description: |
  Introducing the problem of Selective Inference, illustrated through a simple simulation in R.
author:
  - name: vgherard
    url: https://vgherard.github.io
date: 2022-11-14
output:
  distill::distill_article:
    self_contained: false
bibliography: biblio.bib
categories:
  - R
  - Statistics
draft: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prologue

A few months back, for undocumented circumstances, my browser's search
history was full of terms like *"parameter estimation with variable selection"*, 
or *"confidence intervals after cross-validation"*, or again
*"linear model uncertainties after staring into the abyss"*, ... 

Sparing you my rock bottom, I eventually stumbled upon the right keywords, and
started digging into the mathematical aspects of **Selective Inference**, or 
**Post-Model Selection Inference**. Now, while my hands
are still full of dirt, I've decided it's the right moment to write some
notes about what I've learned - whose main recipient is the future me,
which will otherwise inevitably forget what the present me thinks he
knows. If you're *not* the future me:

1. Welcome 👋
2. If you have detected some imprecision, or have suggestions for this or the 
next posts, you are more than welcome to [create an issue](https://github.com/vgherard/vgherard.github.io/issues/new) on the source
repository of this blog.

# Introduction

Broadly speaking, the problem of Selective Inference is that of
performing valid statistical inferences when *the actual questions of the data analysis are not fixed in advance*, but rather *selected through data examination*. In model-based inference, this lack of
pre-determination usually stems from the (often unavoidable) practice of
using the same data to choose an adequate model for the data generating
process *and* to perform inference. The intrinsic
randomness of the selection process has important consequences on the
probability of making different guesses about the selected questions,
which, if not properly taken into account, can completely invalidate the
analysis results.

If this sounds unfamiliar, think about machine-learning: when training a
predictive model on a given dataset, you would usually consider the
error on the same dataset as a poor (optimistic) estimate of the true
model's error rate, because the model was tuned to perform well on that
data in the first place. There we go, Selective Inference! A selection 
from an extended family of models[^1] is performed through data examination,
and this event introduces a bias in the error estimate of the final
model from training data.

[^1]: Here, in the "extended family of models", I'm also implicitly
    accounting for the multiplicity introduced by continuous model
    parameters and training parameters (also known as hyper-parameters).

The example from machine-learning also suggests a very simple-minded and
relatively a-theoretical approach to Selective Inference: data-splitting[^sha]. According to this method, we would use only part of the available data to select the questions to be answered by the analysis, while the remaining part would
be reserved to perform the actual inference. For this program to
succeed, there are however two important requirements: first, we must have 
enough data to ensure decent statistics for both the selection and inference 
tasks; and second, we must be able to split data in two independent 
(or close to independent) sets. This can suppose problems with, *e.g.*, 
time-series data. If, on the other hand, these requirements cannot be met, we 
have to resort to more sophisticated methods.

[^sha]: The preferential method according to [@Shalizi:2020pmsi], from which
I borrowed the "a-theoretical" description, and which I recommend as a starting point for literature review.

At this point, I would like to stress that the conceptual problems
I've just pointed out will probably look obvious to any reader with a
decent intuition for probability[^2]. What is less obvious, but in fact
a [fairly active research field in statistics](https://arxiv.org/search/?query=%22selective+inference%22&searchtype=all&abstracts=show&order=-announced_date_first&size=50),
is how to perform valid selective inferences when the "easy" solution of
data-splitting I mentioned above is not available. This is where theory 
re-enters the game, and what I'm going to ramble about in this and the next 
posts.

[^2]: This is not to say that correctly accounting for Selective Inference is 
the default in scientific practice. A relevant example from the field I come 
from (Particle Physics), is documented in this stimulating reference: 
[@isidori2021significance].

# Illustrations of Selective Inference

Enough for the speech, let us see how selection can affect (and invalidate) 
classical inference with a simple-minded simulation.

### Setting

To illustrate why naive classical inference can fail in the presence of
selection, we consider a very simple regression 
problem involving a single regressor $X$ and a response $Y$, where *all the assumptions of the classical linear model hold*. In fact, we will assume the true data generating process to be:

$$
Y = mX + \varepsilon, \qquad \varepsilon \sim \mathcal N (0, \sigma),
(\#eq:yGaussian)
$$
where $\varepsilon \sim \mathcal N (0, \sigma)$ means "$\varepsilon$ follows a gaussian distribution with mean $0$ and standard deviation $\sigma$".

### A selective modeling procedure

Now, suppose we are given a dataset of $N$ independent observations
$(y_i, x_i)_{i = 1, \,2,\, \dots,\,N}$, and we would like to study the
dependence of $Y$ from $X$. Of course we don't know the true law, Eq.
\@ref(eq:yGaussian), but by a stroke of luck (or by a Taylor expansion
argument) we make the correct initial guess that such dependence is
linear in $X$. We are, however, unsure whether it would be appropriate
to also include an intercept term in the fit. We thus establish the
following selective modeling procedure:

-  Fit a linear model with intercept term,
    $Y = mX + q + \varepsilon$.
-  Stop if the intercept estimate is significantly different from zero (say, at the level of 1-$\sigma$, $p\text{-value}<32\%$). Otherwise:
-  Fit a model with no intercept, $Y = mX + \varepsilon$.

Finally, we *use the last fitted model* to construct a "naive 95%" 
confidence interval $(\hat m_-, \hat m_+)$ for the slope $m$. 
This is defined by:

$$
\hat m_\pm = \hat m\pm t_{0.975, \,N-d} \cdot \hat \sigma _\hat m\qquad (95\%\,\text {C.L.}).
(\#eq:naiveCISlope)
$$ 

Here $t_{0.975,\, N-d}$ is the 97.5%-quantile of the $t$-student
distribution with $N-d$ degrees of freedom, and $d$ is the number of
estimated parameters, ($2$ or $1$, according to where we stopped in the
modeling procedure). $\hat m$ and $\hat \sigma _{\hat m}$ are the
Ordinary Least Squares (OLS) estimates of the slope and its standard
deviation, respectively. These are the classical confidence intervals
reported by the `lm()` function in R.

At a first glance, this procedure might look reasonable. After all, both 
intervals we may end up constructing do have a genuine 95% coverage probability, 
when constructed unconditionally... and by selecting the "best" model we're 
supposedly choosing the "best" confidence interval. In spite of this qualitative 
argument, we inquire:

### ... does it work?

Now, the question is: how often do the naive CIs \@ref(eq:naiveCISlope)
cover the true parameter $m$ of Eq. \@ref(eq:yGaussian)? The answer
better be "at least 95% of the times" for our confidence claim in Eq.
\@ref(eq:naiveCISlope) to be valid!

We can check the actual coverage of \@ref(eq:naiveCISlope) through a simulation. 
Here I'll assume $m = \sigma = 1$, and that the
dataset consists of $N=10$ independent observations of $Y$ at fixed points
$X = (1, \,2, \,\dots ,\, 10)$.

```{r, cache = TRUE}
m <- sigma <- 1  # True parameters
x <- 1:10  # x covariate, assumed fixed
```

The following function generates observations of $Y$ according to the distribution \@ref(eq:yGaussian):

```{r}
generate_y <- function(x, m, sigma) {
	eps <- rnorm(length(x), mean = 0, sd = sigma)
	return(m * x + eps)
	}
```

For example:

```{r}
set.seed(840)
plot(x, generate_y(x, m, sigma), xlab = "X", ylab = "Y")
```

Below we generate $B=10^4$ such $(X,Y)$ datasets, for each of which we fit a linear model according to the procedure specified above, and check how many 
times the true slope $m = 1$ falls in the confidence interval defined by Eq. \@ref(eq:naiveCISlope).

```{r, cache = TRUE}
# Simulation parameters
B <- 1e4  # Number of replications

# Preallocate logical vectors to be assigned for each replica - for efficiency. 
q_dropped <- logical(B)  # Was the intercept term 'q' dropped? 
m_covered <- logical(B)  # Was the true parameter 'm' covered?

# Set seed for reproducibility
set.seed(841)

# Logging
time_start <- Sys.time()  

# Start the simulation
for (b in 1:B) {
	y <- generate_y(x, m, sigma)
	
	# Fit full model (including intercept 'q')
	fit <- lm(y ~ x + 1)  
	q_pval <- summary(fit)$coefficients[1, 4]
	
	# Is 'q' term "significant"? If not, drop 'q' and fit a simpler model
	if (q_pval > 0.32)  { 
		q_dropped[[b]] <- TRUE
		fit <- lm(y ~ x - 1) 
	} else {
		q_dropped[[b]] <- FALSE
	}
	
	# Construct CI for 'm',	using the selected model's fit
	m_ci <- confint(fit, 'x', level = 0.95)
	m_covered[[b]] <- m_ci[[1]] < m && m < m_ci[[2]]
}

time_end <- Sys.time()
cat("Done :) Took ", as.numeric(time_end - time_start), " seconds.")
```

The variable `m_covered[[b]]` is `TRUE` if the slope $m$ fell in the
naive CI $(m_-, m_+)$ defined by Eq. \@ref(eq:naiveCISlope) in the
`b`-th replica of the simulation. Hence, the actual coverage fraction of
the CI is given by:

```{r}
mean(m_covered)  # Actual coverage of naive "95%" CIs.
```

92%! If this difference from the nominal 95% coverage guarantee does not
strike you as enormous, think about it in these terms: the naive CIs
\@ref(eq:naiveCISlope) fail to cover the true parameter about 8% of the
times; This is a relative +60% of failures with respect to an honest 95%
CI.

### What's going on

We can understand a bit better what's happening here by decomposing the
coverage probability as follows:

$$
\text {Pr}(m \in \text{CI})  = \text {Pr}(m \in \text{CI}_{q \text{ dropped}}\,\vert\,q \text{ dropped})\cdot \text {Pr}(q \text{ dropped}) +\\ +\text {Pr}(m \in \text{CI}_{q  \text{ kept}}\,\vert\,q \text{ kept})\cdot \text {Pr}(q \text{ kept})
(\#eq:coverageDecomposition)
$$

The right hand side of this equation shows how our selective modeling
procedure alters the probability $\text{Pr}(m\in \text{CI})$. There are
two contributing factors here: the probability of dropping the intercept
term, and the covering probabilities of the CIs constructed in the two
cases ($\text{CI}_{q \text{ dropped}}$ and
$\text{CI}_{q \text{ kept}}$). We can estimate all these
probabilities as:

```{r}
mean(q_dropped)  # Pr(q dropped)
mean(m_covered[q_dropped])  # Pr(m covered | q dropped)
mean(m_covered[!q_dropped])  # Pr(m covered | q kept)
```

The first result directly follows from our procedure, which uses a
hypothesis test with significance $\alpha = 32\%$ to test the (true)
null hypothesis $q = 0$. It is a bit harder but in fact possible to
prove that[^3]
$\text {Pr}(m \in \text{CI}_{q \text{ dropped}}\,\vert\,q \text{ dropped}) = 95\%$,
as the second estimate would seem to suggest. The third result is
finally what invalidates the naive coverage guarantee in Eq.
\@ref(eq:naiveCISlope).

[^3]: I'm always amazed by the great deal of theory one can learn by
running a dumb simulation, and trying to explain a posteriori what
seems to be a too perfect result. Technically, this follows from the fact that the slope estimate
    $\hat m$ and residual sum of squares $\text{RSS}$ of the reduced
    model, and the $F$-statistic used to test $q = 0$, are all
    independent random variables under the same null hypothesis, here
    true by construction. All these facts are in turn consequences of
    general theorems from linear model theory, see for example
    [@vrbik:2020ra Chapter 4]... and, to be sure, it took me more than a single 
    night without sleep to figure all this out.
    
### Concluding Remarks

To summarize:

1. We started with two linear models for $Y$ vs. $X$, which were in fact *both well-specified* (that is, correct).
2. We stipulated to choose one of the two models by testing the null hypothesis $q = 0$.
3. After selection, we constructed $95\%$ confidence intervals for the slope 
$\hat m$ using the selected model, *as if this had been fixed in advance*.
4. A simulation shows that such intervals have a true coverage probability of
$\approx 92\%$.

The mathematical explanation of the last result is provided by Eq. \@ref(eq:coverageDecomposition), while the (hopefully) plain English one in the introductory part of this post. I will conclude with a few parenthetical remarks. 

First, the selective procedure proposed here would likely hardly be applied in practice in such a simple situation[^4]. However, one could easily think of a more complex scenario with multiple covariates, where eliminating redundant ones could turn out to be beneficial for interpretation (if not compulsory, if the number of covariates exceeds the sample size).

Second, in order to avoid cluttering the discussion with too much 
technicalities, I have deliberately chosen a quite special point in true-model space ($q = 0$). This implies that both fits with and without intercept estimate the *same* slope $m$; this is a peculiar property of $q = 0$, which would not be true in the general case $q \in \mathbb R$. In general, we would have to carefully define the inferential targets for the $q=0$ and $q \in \mathbb R$ cases, in a differential manner.

[^4]: And I'm actually not sure that, after properly taking into account Selective Inference, it would lead to a substantial gain in estimation accuracy, compared to simply fitting the possibly redundant model with intercept.

# Conclusion

That was all for today. In the next post, I will discuss some mathematical details 
regarding the formulation of the Selective Inference problem in model-building. 
For those surviving down to the bottom of the funnel, my future plan is to 
review some (valid) selective inference methods I found interesting, including:

- Benjamini-Yekutieli control of False Coverage Rate [@benjamini2005false],
- POSI bounds for marginal coverage [@berk2013valid],
- Data Fission, an elegant generalization of good old data splitting [@https://doi.org/10.48550/arxiv.2112.11079].
- ...whatever cool stuff I may discover in the meantime.

Ciao!

