---
title: "Automatic resumes of your R-developer portfolio from your R-Universe"
description: |
  Create automatic resumes of your R packages using the R-Universe API.
author:
  - name: Valerio Gherardi
    url: https://vgherard.github.io
date: 2021-07-21
output:
  distill::distill_article:
    self_contained: false
categories:
  - R
draft: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Hi R-bloggers `r emo::ji("wave")` {.unnumbered}

Starting from today, all posts from this blog in the `R` category will also appear on [R-bloggers](https://www.r-bloggers.com/). I would like to thank Tal for aggregating my blog, and say "hi!" to all R-bloggers readers. I'm a particle physicist with a passion for R, Statistics and Machine Learning. If you want to find out something more about me, you can take a look at [my website](https://vgherard.github.io), and links therein.

# Introduction

[R-universe](https://ropensci.org/r-universe/) is a cool initiative from [rOpenSci](https://ropensci.org/), which allows you to create your own CRAN-like repository. The latter is synced with the GitHub repositories (main or specific branches, or releases) associated to your R packages, so that using an R-universe is a very effortless way to organize and share your personal package ecosystem.

If you want to setup your own R-universe, follow the instructions in [this blog post](https://ropensci.org/blog/2021/06/22/setup-runiverse/). In this post, I assume that you have created your own R-universe, and show you how to retrieve metadata on your packages using the R-universe API.

# Retrieving packages descriptions from your R-universe API

Once you will have it set up, your R-universe will be available at the URL `your-user-name.r-universe.dev`. For instance, mine is [vgherard.r-universe.dev](https://vgherard.r-universe.dev). From your R-universe home page, you can access the documentation of the API. We will use the command:

```
GET /stats/descriptions
	NDJSON stream with data from package DESCRIPTION files.
```

The JSON stream can be read with `jsonlite`, as follows:

```{r}
con <- url("https://vgherard.r-universe.dev/stats/descriptions")
pkgs <- jsonlite::stream_in(con)
```

The result is a dataframe with alll the entries of your packages' DESCRIPTION file, e.g.: 

```{r}
pkgs[, c("Package", "Title", "Version")]
```

I use this query on [my personal website](vgherard.github.io/code) to automatically generate a resume of the packages available on my R-universe (this is combined with a GitHub Action scheduled workflow which periodically updates the `Code` section of my website). More precisely, I define an R string `txt` containing the Markdown code for my resume, and I inline it in R Markdown using the synthax `` `r ` ``. This is the code I use on my website:

```{r echo=TRUE}
txt <- ""
for (i in seq_len(nrow(pkgs))) {
	txt <- paste0(
		txt, 
		"### [`", pkgs[i, "Package"], "`](", pkgs[i, "RemoteUrl"], ")", "\n",
		"[![CRAN status](https://www.r-pkg.org/badges/version/", pkgs[i,"Package"],
		")](https://CRAN.R-project.org/package=",pkgs[i, "Package"], ")",
		"\n\n",
		"*", pkgs[i, "Title"], ".* ", pkgs[i, "Description"],
		"\n\n"
		)
}
```

and this is the output:

`r txt`
