---
title: Code
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R packages

The R packages listed below can be installed from [my R-universe public repository](https://vgherard.r-universe.dev/).

*Last update:* `r Sys.time()`

```{r include=FALSE}
con <- url("https://vgherard.r-universe.dev/stats/descriptions")
pkgs <- jsonlite::stream_in(con)

txt <- ""
for (i in seq_len(nrow(pkgs))) {
	pkg_name <- pkgs[i, "Package"]
	pkg_url <- pkgs[i, "RemoteUrl"]
	cran_badge_url <- paste0("https://www.r-pkg.org/badges/version/", pkg_name)
	cran_url <- paste0("https://CRAN.R-project.org/package=", pkg_name)
	
	txt <- paste0(
		txt, 
		"### [`", pkg_name, "`](", pkg_url, ")", " - ",
		"[![CRAN status](", cran_badge_url, ")](", cran_url, ")",
		"\n\n",
		"*", pkgs[i, "Title"], ".* ", pkgs[i, "Description"],
		"\n\n"
		)
}

```

`r txt`

# Miscellaneous

### **[viterbi](https://github.com/vgherard/viterbi)**

*Optimal paths in Hidden Markov Models.* C++ implementation of Viterbi's dynamic programming algorithm for finding optimal paths in the hidden space of hidden Markov models.

### **[dijkstra](https://github.com/vgherard/dijkstra)**

*Minimum length paths in directed graphs.* C++ implementation of Dijkstra's algorithm for finding minimum length paths in weighted directed graphs.


