---
title: Code
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

In my day to day work, I routinely program with [`Python`](https://www.python.org/), SQL, [`R`](https://www.r-project.org/) and [`Bash`](https://tiswww.case.edu/php/chet/bash/bashtop.html). For my personal/fun projects (which is what this page is about 🙂), my preference goes for [`R`](https://www.r-project.org/) and [`C++`](https://www.cplusplus.com/), and 
my main topics of interest are statistical computing, machine-learning and natural language processing.

This page contains a selected list of open-source projects, mostly but not only R packages, I'm working or have worked on.

## R packages

The R packages listed below can be installed from [my R-universe public repository](https://vgherard.r-universe.dev/). This list is automatically updated every day, round midnight (UTC).

*Last update:* `r Sys.time()`

```{r include=FALSE}
con <- url("https://vgherard.r-universe.dev/stats/descriptions")
pkgs <- jsonlite::stream_in(con)

txt <- ""
for (i in seq_len(nrow(pkgs))) {
	txt <- paste0(
		txt, 
		"### [`", pkgs[i, "Package"], "`](", pkgs[i, "RemoteUrl"], ")", "\n",
		"[![CRAN status](https://www.r-pkg.org/badges/version/", pkgs[i,"Package"],
		")](https://CRAN.R-project.org/package=",pkgs[i, "Package"], ")",
		"\n\n",
		"*", pkgs[i, "Title"], ".* ", pkgs[i, "Description"],
		"\n\n"
		)
}

```

`r txt`

# Code excerpts

This Section lists some smaller projects, including code snippets, algorithm implementations and similars.

### **[viterbi](https://github.com/vgherard/viterbi)**

*Optimal paths in Hidden Markov Models.* C++ implementation of Viterbi's dynamic programming algorithm for finding optimal paths in the hidden space of hidden Markov models.

### **[dijkstra](https://github.com/vgherard/dijkstra)**

*Minimum length paths in directed graphs.* C++ implementation of Dijkstra's algorithm for finding minimum length paths in weighted directed graphs.


