---
title: Code
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

My preferred programming languages are [`R`](https://www.r-project.org/) and [`C++`](https://www.cplusplus.com/), and in most of my projects I use a combination of the two (usually via [`{Rcpp}`](http://dirk.eddelbuettel.com/code/rcpp.html)). I also enjoy delving into [`Python`](https://www.python.org/), especially when working with others on machine-learning/deep-learning related projects, either directly or again through its plenty R integrations (*e.g.* [`{reticulate}`](https://rstudio.github.io/reticulate/), [`{tensorflow}`](https://tensorflow.rstudio.com/), [`{keras}`](https://keras.rstudio.com/)). During my research years as physicist, I made extensive use of [`Wolfram`](https://www.wolfram.com/mathematica/) and occasionally worked with [`Fortran`](https://fortran-lang.org/).

This page contains a selected list of open-source projects, mostly but not only R packages, I'm working or have worked on.

## R packages

The R packages listed below can be installed from [my R-universe public repository](https://vgherard.r-universe.dev/).

```{r include=FALSE}
con <- url("https://vgherard.r-universe.dev/stats/descriptions")
pkgs <- jsonlite::stream_in(con)

txt <- ""
for (i in seq_len(nrow(pkgs))) {
	txt <- paste0(
		txt, 
		"### [`", pkgs[i, "Package"], "`](", pkgs[i, "RemoteUrl"], ")", "\n",
		"[![CRAN status](https://www.r-pkg.org/badges/version/", pkgs[i,"Package"],
		")](https://CRAN.R-project.org/package=",pkgs[i, "Package"], ")",
		"\n\n",
		"*", pkgs[i, "Title"], ".* ", pkgs[i, "Description"],
		"\n\n"
		)
}

```

`r txt`

# Code excerpts

This Section lists some smaller projects, including code snippets, algorithm implementations and similars.

### **[viterbi](https://github.com/vgherard/viterbi)**

*Optimal paths in Hidden Markov Models.* C++ implementation of Viterbi's dynamic programming algorithm for finding optimal paths in the hidden space of hidden Markov models.

### **[dijkstra](https://github.com/vgherard/dijkstra)**

*Minimum length paths in directed graphs.* C++ implementation of Dijkstra's algorithm for finding minimum length paths in weighted directed graphs.


