---
title: "Finite sample bias of ARMA estimates"
description: |
  A short description of the post.
author:
  - name: vgherard
    url: https://vgherard.github.io
date: 2023-06-08
output:
  distill::distill_article:
    self_contained: false
draft: true
---

```{r setup, include=FALSE}
set.seed(840)
```


# Intro

Take an AR(1) model such as:

$$
Y_{t+1}= \mu+\alpha (Y_t-\mu) +\epsilon _t. (\#eq:AR1)
$$
The $\alpha$ coefficient estimates returned from functions such as `arima()` 
are *biased*, and the bias can be severe for short time series samples. This is
a simple fact of life of ARMA models (that I learned at my own expense).

Let's illustrate this through a simulation. The block below defines the function 
`sim_fit_ar1()`, which simulates the process of fitting an AR(1) model to actual 
AR(1) data. Concretely, it performs the following two steps:

- Simulate $n$ observations from Eq. \@ref(eq:AR1) above.
- Fit an AR(1) model to the simulated dataset and store the estimated 
coefficients.

These are repeated a large number of times $B$, to obtain an estimate of the 
estimates' distributions.

```{r warning=FALSE, code_folding=TRUE}
sim_ar1 <- function(n, alpha, mu = 0)  
	mu + arima.sim(model = list(order = c(1, 0, 0),  ar1 = alpha), n = n)

fit_ar1 <- function(data, ...) 
	arima(data, order = c(1, 0, 0), ...)

sim_fit_ar1 <- function(B, n, alpha, mu, method = "CSS", plot = TRUE, ...) {
	
	# Simulate from true AR(1) model, fit an AR(1) and extract coefficients
	coefs <- sapply(1:B, \(dummy) {
		sim_ar1(n = n, alpha = alpha, mu = mu) |> 
			fit_ar1(method = method, ...) |> 
			coefficients()
	})
	
	# Remove outliers due to convergence issues using simple heuristic
	# (deviation from median larger than 5 median absolute deviations)
	B_discarded <- 0
	for (i in 1:2) {
		outliers <- abs(coefs[i, ] - median(coefs[i, ])) / mad(coefs[i, ]) > 5
		coefs <- coefs[, !outliers]
		B_discarded <- sum(outliers)
	}
	
	alpha_hat <- coefs["ar1", ]
	mu_hat <- coefs["intercept", ]
	
	if (plot) {
		# Plot histograms of ar1 and intercept coefficients
		oldpar <- par(mfrow=c(1,2))
		
		coef_hist(alpha_hat, alpha, xlab = "alpha")
		coef_hist(mu_hat, mu, xlab = "mu")
		title(main = "Distribution of AR(1) model estimates",
					sub = paste0("B: ", B, " - n: ", n, " - method: ", method, 
											 " - discarded ", B_discarded, " simulations"),
					outer = TRUE,
					line = -1
					)
		par(oldpar)
	}
	
	# Return summary of results
	res <- data.frame(
		n = n, 
		method = method, 
		alpha_hat_mean = mean(alpha_hat),
		alpha_hat_se = sd(alpha_hat) / sqrt(B),
		alpha = alpha,
		mu_hat_mean = mean(mu_hat),
		mu_hat_se = sd(mu_hat) / sqrt(B),
		mu = mu,
		B = B,
		B_discarded = B_discarded
	)
	return(invisible(res))
}

coef_hist <- function(estimates, target, xlab) {
	hist(estimates, freq = FALSE, xlab = xlab, main = NULL)
	abline(v = mean(estimates), lty = "dashed", lwd = 4)
	abline(v = target, col = "red", lty = "dashed", lwd = 4)
}
```


# Simulations


## Warm up

We start by comparing the distributions of $\hat \alpha$ for 
$n = 10, \,100,\,1000$, for fixed $\alpha = 0.5$, $\mu = 1$. I've silently
used the default (in `arima.sim()`) variance $\sigma ^2 =1$ for the innovations 
$\epsilon _t$, so that $\mu$ can be interpreted here as a sort of signal to 
noise ratio.

We see that $\hat \alpha$ has a relative bias of about 50% (!) for $n=10$, which
gradually decreases as $n$ increases (since $\hat \alpha$ is a consistent 
estimator). 

The plot for $n = 10$ reports that $110$ simulations (roughly 1% of the 
total $B = 10^4$ simulations) were discarded from the reported results. This is
because such simulations had convergence issues, and I manually removed these 
results, a problem that we don't really need to worry too much about
(not with a 50% bias!).


```{r warning=FALSE}
set.seed(840)
sim_fit_ar1(10000, 10, alpha = 0.5, mu = 1, method = "CSS")
```

```{r}
sim_fit_ar1(10000, 100, alpha = 0.5, mu = 1, method = "CSS")
```

```{r}
sim_fit_ar1(10000, 1000, alpha = 0.5, mu = 1, method = "CSS")
```

## Parameter space scan

Scan in:
- `n`, `method`, `mu`, `alpha`
