---
title: "kgrams v0.1.2 on CRAN"
description: |
  kgrams: Classical k-gram Language Models in R.
date: 2021-11-13
categories:
  - Natural Language Processing
  - R
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

Version v0.1.2 of my R package [kgrams](https://github.com/vgherard/kgrams) was just accepted by CRAN. This package provides tools for training and evaluating k-gram language models in R, supporting several probability smoothing techniques, perplexity computations, random text generation and more.

## Short demo

```{r}
library(kgrams)
# Get k-gram frequency counts from Shakespeare's "Much Ado About Nothing"
freqs <- kgram_freqs(kgrams::much_ado, N = 4)

# Build modified Kneser-Ney 4-gram model, with discount parameters D1, D2, D3.
mkn <- language_model(freqs, smoother = "mkn", D1 = 0.25, D2 = 0.5, D3 = 0.75)

# Sample sentences from the language model at different temperatures
set.seed(840)
sample_sentences(model = mkn, n = 3, max_length = 10, t = 1)
sample_sentences(model = mkn, n = 3, max_length = 10, t = 0.1)
sample_sentences(model = mkn, n = 3, max_length = 10, t = 10)
```

## NEWS

### Overall Software Improvements

-   The package's test suite has been greatly extended.
-   Improved error/warning conditions for wrong arguments.
-   Re-enabled compiler diagnostics as per CRAN policy (#19)

### API Changes

-   `verbose` arguments now default to `FALSE`.
-   `probability()`, `perplexity()` and `sample_sentences()` are restricted to accept only `language_model` class objects as their `model` argument.

### New features

-   `as_dictionary(NULL)` now returns an empty `dictionary`.

### Bug Fixes

-   Fixed bug causing `.preprocess` and `.tknz_sent` arguments to be ignored in `process_sentences()`.
-   Fixed previously wrong defaults for `max_lines` and `batch_size` arguments in `kgram_freqs.connection()`.
-   Added print method for class `dictionary`.
-   Fixed bug causing invalid results in `dictionary()` with batch processing and non-trivial size constraints on vocabulary size.

### Other

-   Maintainer's email updated
